Finding shortest and closest vectors in a lattice of Voronoi's first kind

### 2015-07-02 talk/presentation.tex
minor mods to talk

Longstanding problem in number theory and computer science:
\begin{itemize}
\item sphere packing {\tiny (Voronoi 1908, Conway and Sloane 1999)}
\item integer programming {\tiny (Lenstra~et.~al. 1983, Kannan 1987)}
\item the factoring of polynomials over rationals {\tiny (Lenstra~et.~al. 1982)}
\item checking solvability by radicals {\tiny (Landau and Miller 1985)}
\item solving low density subset sum problems {\tiny (Coster~et.~al. 1992)}
\end{itemize}
}

@@

Communications applications:
\begin{itemize}
\item minimum distance of lattice codes {\tiny (Conway and Sloane 1982, Erez~et.~al. 2004)}
\item capacity analysis of network codes {\tiny (Nazer and Gastpar 2011)}
\end{itemize}
}

\frame{
\frametitle{The shortest vector problem}
Central problem in modern cryptography enabling:
\begin{itemize}
\item efficient public key encryption {\tiny (Gentry~et.~al. 2008, Micciancio 2012)}
\item post quantum cryptography {\tiny (Perlner~et.~al. 2009, Guneysu 2014, Peikert 2014)}
%\item efficient hash functions {\tiny (Peikert and Rosen 2006, Micciancio 2007)}
\item fully homomorphic encryption {\tiny (Gentry~et.~al. 2009, 2010)}
\end{itemize}

@@

\frame{
\frametitle{The shortest vector problem}
\begin{itemize}
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item Short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind} {\tiny (Voronoi~1908, Conway~and~Sloane~1992, McKilliam~and~Grant~2012)}.
\end{itemize}
}

@@

\frame{
\frametitle{The closest lattice point problem}
Communications applications:
\begin{itemize}
\item spectrally efficient codes {\tiny (Conway and Sloane 1983, Erez~et.~al.~2004)},
\item quantisation and compression {\tiny (Conway~and~Sloane~1982, Erez~et.~al.~2005)},
\item multi-antenna communication (MIMO) {\tiny (El Gamal~et.~al.~2004, Wubben~et.~al.~2011)},
\item equalisation of delay spread channels {\tiny (Zamir~et.~al.~2002, Damen~et.~al.~2003)},
\item network coding {\tiny (Nazer and Gastpar 2011)},
\item blind channel estimation {\tiny (Ryan~et.~al. 2007, McKilliam~2014)}.
\end{itemize}

@@

\frame{
\frametitle{The closest lattice point problem}
Signal processing applications:
\begin{itemize}
\item unwrapping of phase data for electronic distance measurement in GPS and surveying {\tiny (Tuenissen~1995, Hassibi~and~Boyd~1998)},
\item single frequency estimation {\tiny (McKilliam~et.~al.~2010)},
\item polynomial phase estimation {\tiny (McKilliam~et.~al.~2014)},
\item circular statistics {\tiny (McKilliam~et.~al.~2014)}.
\end{itemize}

@@

@@

\frame{
\frametitle{A series of relevant vectors}
\begin{itemize}
\item The number of iterations depends on $x_0$ and might be large.
\item Minimising over the set of relevant vectors, that is computing
\[
\arg\min_{ v \in \relevant(\Lambda) \cup \{0\} } \|y - x_k - v \|
\]
might be expensive.
\item There are as many as $2^{n+1}-2$ relevant vectors.
\end{itemize}
}

>>>>

Longstanding problem in number theory and computer science:
\begin{itemize}
\item sphere packing %{\tiny (Voronoi 1908, Conway and Sloane 1999)}
\item integer programming %{\tiny (Lenstra~et.~al. 1983, Kannan 1987)}
\item the factoring of polynomials over rationals %{\tiny (Lenstra~et.~al. 1982)}
\item checking solvability by radicals %{\tiny (Landau and Miller 1985)}
\item solving low density subset sum problems %{\tiny (Coster~et.~al. 1992)}
\end{itemize}
}

@@

Communications applications:
\begin{itemize}
\item minimum distance of lattice codes %{\tiny (Conway and Sloane 1982, Erez~et.~al. 2004)}
\item capacity analysis of network codes %{\tiny (Nazer and Gastpar 2011)}
\end{itemize}
}

\frame{
\frametitle{The shortest vector problem}
Central problem in modern cryptography enabling:
\begin{itemize}
\item efficient public key encryption %{\tiny (Gentry~et.~al. 2008, Micciancio 2012)}
\item post quantum cryptography %{\tiny (Perlner~et.~al. 2009, Guneysu 2014, Peikert 2014)}
%\item efficient hash functions {\tiny (Peikert and Rosen 2006, Micciancio 2007)}
\item fully homomorphic encryption %{\tiny (Gentry~et.~al. 2009, 2010)}
\end{itemize}

@@

\frame{
\frametitle{The shortest vector problem}
\begin{itemize}
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item Short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$, and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind} %{\tiny (Voronoi~1908, Conway~and~Sloane~1992, McKilliam~and~Grant~2012)}.
\end{itemize}
}

@@

\frame{
\frametitle{The closest lattice point problem}
Communications applications:
\begin{itemize}
\item spectrally efficient codes %{\tiny (Conway and Sloane 1983, Erez~et.~al.~2004)},
\item quantisation and compression %{\tiny (Conway~and~Sloane~1982, Erez~et.~al.~2005)},
\item multi-antenna communication (MIMO) %{\tiny (El Gamal~et.~al.~2004, Wubben~et.~al.~2011)},
\item equalisation of delay spread channels %{\tiny (Zamir~et.~al.~2002, Damen~et.~al.~2003)},
\item network coding %{\tiny (Nazer and Gastpar 2011)},
\item blind channel estimation % {\tiny (Ryan~et.~al. 2007, McKilliam~2014)}.
\end{itemize}

@@

\frame{
\frametitle{The closest lattice point problem}
Signal processing applications:
\begin{itemize}
\item unwrapping of phase data for electronic distance measurement in GPS and surveying %{\tiny (Tuenissen~1995, Hassibi~and~Boyd~1998)},
\item single frequency estimation %{\tiny (McKilliam~et.~al.~2010)},
\item polynomial phase estimation %{\tiny (McKilliam~et.~al.~2014)},
\item circular statistics %{\tiny (McKilliam~et.~al.~2014)}.
\end{itemize}

@@

% \frame{
% \frametitle{The closest lattice point problem}
% Central problem in modern cryptography enabling:
% \begin{itemize}
% \item efficient public key encryption %{\tiny (Gentry~et.~al. 2008, Micciancio 2012)}
% \item post quantum cryptography %{\tiny (Perlner~et.~al. 2009, Guneysu 2014, Peikert 2014)}
% %\item efficient hash functions {\tiny (Peikert and Rosen 2006, Micciancio 2007)}
% \item fully homomorphic encryption %{\tiny (Gentry~et.~al. 2009, 2010)}
% \end{itemize}

% Security depends upon the complexity of the closest lattice point vector problem.
% }

@@

\frame{
\frametitle{A series of relevant vectors}
\begin{itemize}
\item The number of iterations depends on $x_0$ and might be large.
\item Minimising over the set of relevant vectors, that is, computing
\[
\arg\min_{ v \in \relevant(\Lambda) \cup \{0\} } \|y - x_k - v \|
\]
might be expensive.
\item There are as many as $2^{n+1}-2$ relevant vectors.
\end{itemize}
}
